---
title: "Homework 5: ISYE 6501 - Introduction to Analytics Modeling"
output: pdf_document
bibliography: docs/references.bib
csl: docs/ieee.csl
urlcolor: blue
# date: "2024-09-26"
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

# Question 9.1

### Prompt

Using the same crime data set `uscrime.txt` as in Question 8.2, apply 
Principal Component Analysis and then create a regression model using the 
first few principal components.  
Specify your new model in terms of the original variables 
(not the principal components), and compare its quality to that of your 
solution to Question 8.2.  
You can use the R function `prcomp` for PCA. 
(**Note** that to first scale the data, you can include `scale = TRUE` to scale
as part of the PCA function. 
Don’t forget that, to make a prediction for the new city, you’ll need to 
unscale the coefficients (i.e., do the scaling calculation in reverse)!)

### Solution

- Apply Principal Component Analysis (PCA)
- Create regression model using first few principal components
- Specify model in terms of *original variables* not principle components
- Compare quality to solution to Question 8.2
  - 15 predictors, 6 predictors (higher adjusted R-squared)
- Use `prcomp()` for PCA
- Scale data for PCA function
- Make prediction *without scaling*

### Initial Approach

```{r, message=FALSE}
# Import libraries
# lm and glm are in the stats package loaded by default
library(ggplot2) # Plot functions

# Set the working directory
setwd("~/projects/ISYE6501/HW5")

# Load the crime data into a table
data <- read.table("data/uscrime.txt", stringsAsFactors = FALSE, header = TRUE)

# View the imported data to check import but only first few rows
head(data, 5)
```


