---
title: "Homework 8: ISYE 6501 - Introduction to Analytics Modeling"
output: pdf_document
bibliography: docs/references.bib
csl: docs/ieee.csl
urlcolor: blue
# date: "2024-10-02"
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

# Question 11.1

### Prompt

Using the crime data set `uscrime.txt` from Questions 8.2, 9.1, and 10.1, 
build a regression model using:

1.	Stepwise regression
2.	Lasso
3.	Elastic net

For Parts 2 and 3, remember to scale the data first – otherwise, the 
regression coefficients will be on different scales and the constraint won’t 
have the desired effect.

For Parts 2 and 3, use the `glmnet` function in R.  

Notes on R:

- For the elastic net model, what we called λ in the videos, `glmnet` calls 
“alpha”; you can get a range of results by varying alpha from 1 (lasso) to 
0 (ridge regression) [and, of course, other values of alpha in between].
- In a function call like `glmnet(x,y,family=”mgaussian”,alpha=1)` the 
predictors x need to be in R’s matrix format, rather than data frame format.  
You can convert a data frame to a matrix using `as.matrix` – for example, 
`x <- as.matrix(data[,1:n-1])`
- Rather than specifying a value of T, `glmnet` returns models for a variety of 
values of T.


### Solution


For part 2 and 3, `glmnet` is used to fit a generalized linear model
with lasso or elastic net regularization [@glmnetFunction].


### Stepwise Regression

Stepwise regression is a combination of forward selection and backward
elimination.

```{r, message=FALSE}
# Load packages

# Set seed so results are reproducible
set.seed(123)

# Set the working directory
setwd("~/projects/ISYE6501/HW8")

# Load the US crime data data into a table
data <- read.table("data/uscrime.txt", stringsAsFactors = FALSE, header = TRUE)
cat(paste("\nUS Crime Data:\n"))
print(head(data, 5))
```

268

### Lasso

The LASSO method adds a constraint to the standard regression equation.
Lasso regression is utilized in an effort to choose coefficients $a_1,\dots,a_n$
to minimize total error defined by:

$$
  \sum_{i=1}^m \bigg(y_i - (a_0+\sum^n_{j=1}a_j x_{ij})\bigg)^2
$$
Lasso regression is restricted to solutions where:

$$
  \sum_{j=1}^n |a_j| \leq \tau_{Lasso}
$$

The data must be scaled.

### Elastic Net

Elastic net constrains a combination of absolute value of coefficients
and their squares.

The data must be scaled.



\newpage
# References
